---
title: "ShrinkageTrees: Bayesian Tree Ensembles for Survival Analysis and Causal Inference"
author: "Tijn Jacobs"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{ShrinkageTrees: Introduction and Usage}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
set.seed(42)
```

## Introduction

**ShrinkageTrees** is an R package for Bayesian regression tree ensembles
designed for two challenging settings in modern statistics:

- **High-dimensional survival analysis** — right-censored outcomes with many
  more covariates than observations.
- **Causal inference with heterogeneous treatment effects** — estimating
  patient-level conditional average treatment effects (CATEs) and the
  population-level average treatment effect (ATE).

The package builds on the well-known **BART** framework
(Chipman, George & McCulloch 2010) and extends it in three directions:

1. **Horseshoe leaf shrinkage.** A global–local horseshoe prior is placed
   directly on the step heights (leaf parameters) of each tree, allowing
   strong signals to retain their magnitude while noise is aggressively
   shrunk toward zero.

2. **AFT survival models.** Right-censored survival times are handled through
   an accelerated failure time (AFT) formulation with Bayesian data
   augmentation, enabling full posterior inference without
   proportional-hazards assumptions.

3. **BCF-style causal decomposition.** Following Hahn, Murray & Carvalho
   (2020), the outcome is decomposed into a prognostic function
   $\mu(\mathbf{x})$ and a treatment-effect function $\tau(\mathbf{x})$,
   each estimated by a separate tree ensemble.

### Package map

| Function | Task | Prior |
|---|---|---|
| `HorseTrees()` | Prediction (continuous / binary / survival) | Horseshoe |
| `ShrinkageTrees()` | Prediction — flexible prior choice | Horseshoe, DART, BART, … |
| `SurvivalBART()` | Survival prediction | Classical BART |
| `SurvivalDART()` | Sparse survival prediction | DART (Dirichlet) |
| `SurvivalBCF()` | Causal survival inference | BCF (classical) |
| `SurvivalShrinkageBCF()` | Sparse causal survival inference | BCF + DART |
| `CausalHorseForest()` | Causal inference (all outcomes) | Horseshoe |
| `CausalShrinkageForest()` | Causal inference — flexible prior | Horseshoe, DART, BART, … |

All model-fitting functions return an S3 object with consistent
`print()`, `summary()`, `predict()`, and `plot()` methods.

---

## The PDAC Dataset

The package ships with `pdac`, a processed subset of the TCGA pancreatic
ductal adenocarcinoma (PAAD) cohort. The dataset contains overall survival
times, a binary treatment indicator (radiation therapy vs. control), clinical
covariates, and expression values of approximately 3,000 genes selected by
median absolute deviation.

```{r load-data}
library(ShrinkageTrees)
data("pdac")

# Dimensions and column overview
cat("Patients:", nrow(pdac), "\n")
cat("Columns :", ncol(pdac), "\n")
cat("Clinical columns:", paste(names(pdac)[1:13], collapse = ", "), "\n")
cat("Survival: time (months), censoring rate =",
    round(1 - mean(pdac$status), 2), "\n")
cat("Treatment: radiation =", sum(pdac$treatment),
    "/ control =", sum(1 - pdac$treatment), "\n")
```

We separate the outcome, treatment, and covariate matrix for the analyses
below.

```{r prepare-data}
time      <- pdac$time
status    <- pdac$status
treatment <- pdac$treatment
X         <- as.matrix(pdac[, !(names(pdac) %in% c("time", "status", "treatment"))])
```

---

## Single-Forest Models

### HorseTrees — binary outcome (propensity scores)

Before fitting causal models we estimate propensity scores
$\hat{e}(\mathbf{x}) = P(W=1 \mid \mathbf{x})$ using `HorseTrees()` with a
binary outcome. The probit link is used internally: predictions are on the
latent Gaussian scale and can be converted to probabilities with `pnorm()`.

The code block below uses reduced MCMC settings for illustration.
A real analysis would use `N_post = 5000, N_burn = 5000`.

```{r horsetrees-binary, eval=FALSE}
ps_fit <- HorseTrees(
  y            = treatment,
  X_train      = X,
  outcome_type = "binary",
  k            = 0.1,
  N_post       = 5000,
  N_burn       = 5000,
  verbose      = FALSE
)

propensity <- pnorm(ps_fit$train_predictions)
```

For the remainder of this vignette we use a short synthetic run to keep
build time low.

```{r horsetrees-binary-small}
set.seed(1)
n <- 80;  p <- 10
X_syn  <- matrix(rnorm(n * p), n, p)
W_syn  <- rbinom(n, 1, pnorm(0.8 * X_syn[, 1]))

ps_fit <- HorseTrees(
  y            = W_syn,
  X_train      = X_syn,
  outcome_type = "binary",
  number_of_trees = 5,
  k            = 0.5,
  N_post       = 50,
  N_burn       = 25,
  verbose      = FALSE
)

propensity_syn <- pnorm(ps_fit$train_predictions)
cat("Propensity scores — range: [",
    round(range(propensity_syn), 3), "]\n")
```

---

### HorseTrees — survival outcome

`HorseTrees()` handles right-censored data via an AFT model.
Pass `outcome_type = "right-censored"` and provide the `status` vector.

```{r horsetrees-survival}
set.seed(2)
log_T <- X_syn[, 1] + rnorm(n)
C     <- rexp(n, 0.5)
y_syn   <- pmin(exp(log_T), C)
d_syn   <- as.integer(exp(log_T) <= C)

ht_surv <- HorseTrees(
  y               = y_syn,
  status          = d_syn,
  X_train         = X_syn,
  outcome_type    = "right-censored",
  timescale       = "time",
  number_of_trees = 5,
  N_post          = 50,
  N_burn          = 25,
  store_posterior_sample = TRUE,
  verbose         = FALSE
)

cat("Posterior mean log-time (first 5 obs):",
    round(ht_surv$train_predictions[1:5], 3), "\n")
cat("Posterior sigma — mean:",
    round(mean(ht_surv$sigma), 3), "\n")
```

---

### ShrinkageTrees — flexible prior choice

`ShrinkageTrees()` generalises `HorseTrees()` by exposing the `prior_type`
argument. Available priors are `"horseshoe"`, `"horseshoe_fw"`,
`"horseshoe_EB"`, `"half-cauchy"`, `"standard"` (BART), and `"dirichlet"`
(DART).

```{r shrinkage-continuous}
set.seed(3)
y_cont <- X_syn[, 1] + 0.5 * X_syn[, 2] + rnorm(n)

# Horseshoe prior (default for HorseTrees)
fit_hs <- ShrinkageTrees(
  y               = y_cont,
  X_train         = X_syn,
  outcome_type    = "continuous",
  prior_type      = "horseshoe",
  local_hp        = 0.1 / sqrt(5),
  global_hp       = 0.1 / sqrt(5),
  number_of_trees = 5,
  N_post          = 50,
  N_burn          = 25,
  verbose         = FALSE
)

# Forest-wide horseshoe (horseshoe_fw)
fit_fw <- ShrinkageTrees(
  y               = y_cont,
  X_train         = X_syn,
  outcome_type    = "continuous",
  prior_type      = "horseshoe_fw",
  local_hp        = 0.1 / sqrt(5),
  global_hp       = 0.1 / sqrt(5),
  number_of_trees = 5,
  N_post          = 50,
  N_burn          = 25,
  verbose         = FALSE
)

cat("Horseshoe   — train RMSE:",
    round(sqrt(mean((fit_hs$train_predictions - y_cont)^2)), 3), "\n")
cat("Horseshoe FW— train RMSE:",
    round(sqrt(mean((fit_fw$train_predictions - y_cont)^2)), 3), "\n")
```

---

### Survival wrappers: SurvivalBART and SurvivalDART

The wrappers `SurvivalBART()` and `SurvivalDART()` provide convenience
interfaces for classical BART and DART under the AFT formulation.
They calibrate prior hyperparameters automatically from the data range.

```{r survival-wrappers}
set.seed(4)

# SurvivalBART: classical BART prior, AFT likelihood
fit_sbart <- SurvivalBART(
  time            = y_syn,
  status          = d_syn,
  X_train         = X_syn,
  number_of_trees = 5,
  k               = 2.0,
  N_post          = 50,
  N_burn          = 25,
  verbose         = FALSE
)

# SurvivalDART: Dirichlet (DART) splitting prior
fit_sdart <- SurvivalDART(
  time            = y_syn,
  status          = d_syn,
  X_train         = X_syn,
  number_of_trees = 5,
  k               = 2.0,
  N_post          = 50,
  N_burn          = 25,
  verbose         = FALSE
)

cat("SurvivalBART  class:", class(fit_sbart), "\n")
cat("SurvivalDART  class:", class(fit_sdart), "\n")
```

---

## High-Dimensional Survival Analysis

A key motivation for horseshoe shrinkage and the Dirichlet (DART) sparsity
prior is their behaviour in the **$p \gg n$ regime**: many covariates are
available but only a small subset drives the outcome. Classical BART may
struggle here because the standard Gaussian leaf prior is non-sparse and
does not concentrate on a small number of predictors.

We illustrate both priors on a sparse AFT simulation: $n = 60$ observations,
$p = 200$ predictors, and only three active predictors.

```{r hd-data}
set.seed(20)
n_hd <- 60;  p_hd <- 200
X_hd <- matrix(rnorm(n_hd * p_hd), n_hd, p_hd)

# True log-survival depends only on predictors 1, 2, and 3
log_T_hd <- 1.5 * X_hd[, 1] - 1.0 * X_hd[, 2] + 0.5 * X_hd[, 3] + rnorm(n_hd)
C_hd     <- rexp(n_hd, rate = 0.5)
y_hd     <- pmin(exp(log_T_hd), C_hd)
d_hd     <- as.integer(exp(log_T_hd) <= C_hd)

cat("n =", n_hd, "| p =", p_hd,
    "| active predictors = 3",
    "| censoring rate =", round(1 - mean(d_hd), 2), "\n")
```

**ShrinkageTrees (horseshoe)** places global–local shrinkage on the step
heights of every leaf, automatically regularising all 200 predictors
toward zero while preserving the signal in the three active ones.

```{r hd-horseshoe}
set.seed(21)
fit_hd_hs <- ShrinkageTrees(
  y               = y_hd,
  status          = d_hd,
  X_train         = X_hd,
  outcome_type    = "right-censored",
  prior_type      = "horseshoe",
  local_hp        = 0.1 / sqrt(10),
  global_hp       = 0.1 / sqrt(10),
  number_of_trees = 10,
  N_post          = 50,
  N_burn          = 25,
  verbose         = FALSE
)
```

**SurvivalDART** uses a Dirichlet prior on split probabilities to induce
structural sparsity: after burn-in, most splitting probability is
concentrated on truly predictive variables. Setting `rho_dirichlet = 3`
encodes the prior belief that approximately three predictors are active.

```{r hd-dart}
set.seed(22)
fit_hd_dart <- SurvivalDART(
  time            = y_hd,
  status          = d_hd,
  X_train         = X_hd,
  number_of_trees = 10,
  rho_dirichlet   = 3,
  N_post          = 50,
  N_burn          = 25,
  verbose         = FALSE
)
```

Both models run without error in the $p > n$ regime. We compare their
posterior mean predictions in log-time against the latent true values
used to generate the data.

```{r hd-compare}
rmse_hs   <- sqrt(mean((fit_hd_hs$train_predictions  - log_T_hd)^2))
rmse_dart <- sqrt(mean((fit_hd_dart$train_predictions - log_T_hd)^2))

cat(sprintf("%-18s  train RMSE (log-time): %.3f\n", "Horseshoe", rmse_hs))
cat(sprintf("%-18s  train RMSE (log-time): %.3f\n", "DART",      rmse_dart))
```

The DART model also produces **variable importance** plots that display
the posterior distribution of each predictor's splitting probability.
With only 50 posterior draws the top-10 plot below should already concentrate
most probability mass near the three truly active predictors.

```{r hd-vi, eval=requireNamespace("bayesplot", quietly=TRUE) && requireNamespace("ggplot2", quietly=TRUE)}
plot(fit_hd_dart, type = "vi", n_vi = 10)
```

---

## Causal Forest Models

The causal models decompose the outcome as
$$
Y_i = \mu(\mathbf{x}_i) + W_i \cdot \tau(\mathbf{x}_i) + \varepsilon_i,
$$
where $\mu(\cdot)$ is the prognostic (control) function and $\tau(\cdot)$ is
the heterogeneous treatment effect. Separate tree ensembles are fitted for
each component.

We generate synthetic data with a known treatment effect for illustration.

```{r causal-data}
set.seed(5)
tau_true <- 1.5 * (X_syn[, 1] > 0)    # heterogeneous treatment effect
y_causal <- X_syn[, 1] + W_syn * tau_true + rnorm(n, sd = 0.5)
```

---

### SurvivalBCF — classical BCF for survival

`SurvivalBCF()` fits a BCF model for right-censored survival outcomes using
classical BART priors.

```{r survbcf, eval=FALSE}
# Full analysis (eval=FALSE — use larger MCMC settings in practice)
fit_sbcf <- SurvivalBCF(
  time       = time,
  status     = status,
  X_train    = X,
  treatment  = treatment,
  propensity = propensity,   # from HorseTrees above
  N_post     = 5000,
  N_burn     = 5000,
  verbose    = FALSE
)
```

```{r survbcf-small}
set.seed(6)
fit_sbcf <- SurvivalBCF(
  time            = y_syn,
  status          = d_syn,
  X_train         = X_syn,
  treatment       = W_syn,
  number_of_trees_control = 5,
  number_of_trees_treat   = 5,
  N_post          = 50,
  N_burn          = 25,
  verbose         = FALSE
)
cat("SurvivalBCF class:", class(fit_sbcf), "\n")
cat("ATE (posterior mean):",
    round(mean(fit_sbcf$train_predictions_treat), 3), "\n")
```

---

### SurvivalShrinkageBCF — sparse causal survival forest

`SurvivalShrinkageBCF()` extends BCF with a Dirichlet splitting prior on
both forests, inducing sparsity in high-dimensional settings.

```{r survsbcf-small}
set.seed(7)
fit_ssbcf <- SurvivalShrinkageBCF(
  time            = y_syn,
  status          = d_syn,
  X_train         = X_syn,
  treatment       = W_syn,
  number_of_trees_control = 5,
  number_of_trees_treat   = 5,
  N_post          = 50,
  N_burn          = 25,
  verbose         = FALSE
)
cat("SurvivalShrinkageBCF class:", class(fit_ssbcf), "\n")
```

---

### CausalHorseForest — horseshoe causal forest

`CausalHorseForest()` is the primary novel contribution of this package.
It applies horseshoe shrinkage to the leaf parameters of both the prognostic
and treatment-effect forests. This enables effective regularisation when
many covariates are available but few are truly predictive of heterogeneous
treatment effects.

```{r causal-horse}
set.seed(8)
fit_chf <- CausalHorseForest(
  y                         = y_causal,
  X_train_control           = X_syn,
  X_train_treat             = X_syn,
  treatment_indicator_train = W_syn,
  outcome_type              = "continuous",
  number_of_trees           = 5,
  N_post                    = 50,
  N_burn                    = 25,
  store_posterior_sample    = TRUE,
  verbose                   = FALSE
)

cat("CausalHorseForest class:", class(fit_chf), "\n")

# Posterior mean CATE
cate_mean <- fit_chf$train_predictions_treat
cat("CATE — posterior mean (first 5):",
    round(cate_mean[1:5], 3), "\n")

# Posterior ATE
ate_samples <- rowMeans(fit_chf$train_predictions_sample_treat)
cat("ATE posterior mean:",
    round(mean(ate_samples), 3),
    "  95% CI: [",
    round(quantile(ate_samples, 0.025), 3), ",",
    round(quantile(ate_samples, 0.975), 3), "]\n")
```

You can also supply separate test matrices to obtain out-of-sample predictions.

```{r causal-horse-test}
set.seed(9)
X_test <- matrix(rnorm(20 * p), 20, p)
W_test <- rbinom(20, 1, 0.5)

fit_chf_test <- CausalHorseForest(
  y                         = y_causal,
  X_train_control           = X_syn,
  X_train_treat             = X_syn,
  treatment_indicator_train = W_syn,
  X_test_control            = X_test,
  X_test_treat              = X_test,
  treatment_indicator_test  = W_test,
  outcome_type              = "continuous",
  number_of_trees           = 5,
  N_post                    = 50,
  N_burn                    = 25,
  store_posterior_sample    = TRUE,
  verbose                   = FALSE
)

cat("Test CATE (first 5):",
    round(fit_chf_test$test_predictions_treat[1:5], 3), "\n")
```

---

### CausalShrinkageForest — flexible causal priors

`CausalShrinkageForest()` is the most general interface. It allows
independent prior choices for the control and treatment forests via
`prior_type_control` and `prior_type_treat`.

```{r causal-shrinkage}
set.seed(10)
lh <- 0.1 / sqrt(5)

fit_csf <- CausalShrinkageForest(
  y                         = y_causal,
  X_train_control           = X_syn,
  X_train_treat             = X_syn,
  treatment_indicator_train = W_syn,
  outcome_type              = "continuous",
  prior_type_control        = "horseshoe",
  prior_type_treat          = "horseshoe",
  local_hp_control          = lh,
  global_hp_control         = lh,
  local_hp_treat            = lh,
  global_hp_treat           = lh,
  number_of_trees_control   = 5,
  number_of_trees_treat     = 5,
  N_post                    = 50,
  N_burn                    = 25,
  store_posterior_sample    = TRUE,
  verbose                   = FALSE
)

cat("CausalShrinkageForest class:", class(fit_csf), "\n")
cat("Acceptance ratio (control):",
    round(fit_csf$acceptance_ratio_control, 3), "\n")
cat("Acceptance ratio (treat)  :",
    round(fit_csf$acceptance_ratio_treat, 3), "\n")
```

The `horseshoe_fw` prior adds a forest-wide shrinkage parameter that is
tracked in the fitted object.

```{r causal-shrinkage-fw}
set.seed(11)
fit_fw2 <- CausalShrinkageForest(
  y                         = y_causal,
  X_train_control           = X_syn,
  X_train_treat             = X_syn,
  treatment_indicator_train = W_syn,
  outcome_type              = "continuous",
  prior_type_control        = "horseshoe_fw",
  prior_type_treat          = "horseshoe_fw",
  local_hp_control          = lh,
  global_hp_control         = lh,
  local_hp_treat            = lh,
  global_hp_treat           = lh,
  number_of_trees_control   = 5,
  number_of_trees_treat     = 5,
  N_post                    = 50,
  N_burn                    = 25,
  verbose                   = FALSE
)

cat("Forest-wide shrinkage (control, first 5 draws):\n")
print(round(fit_fw2$forestwide_shrinkage_control[1:5], 4))
```

---

## S3 Methods

All fitted objects support `print()`, `summary()`, `predict()`, and
`plot()`.

### print()

Calling `print()` (or just typing the object name) displays a concise model
summary.

```{r print}
print(fit_chf)
```

For causal models the output additionally shows the number of trees in each
forest and prior details for both components.

```{r print-csf}
print(fit_csf)
```

---

### summary()

`summary()` returns a structured list and displays a richer description
including posterior statistics for $\sigma$, acceptance ratios, and
treatment effect estimates.

```{r summary-shrinkage}
smry <- summary(fit_hs)
print(smry)
```

For causal models the summary includes the posterior ATE with a 95% credible
interval (when `store_posterior_sample = TRUE`).

```{r summary-causal}
smry_c <- summary(fit_chf)
print(smry_c)

# Access the ATE directly
cat("ATE mean  :", round(smry_c$treatment_effect$ate, 3), "\n")
cat("ATE 95% CI: [",
    round(smry_c$treatment_effect$ate_lower, 3), ",",
    round(smry_c$treatment_effect$ate_upper, 3), "]\n")
```

---

### predict()

`predict()` computes the posterior predictive distribution on new data.
It returns a `ShrinkageTreesPrediction` object with posterior mean and
credible-interval vectors.

```{r predict}
X_new  <- matrix(rnorm(10 * p), 10, p)

pred <- predict(fit_hs, newdata = X_new)
print(pred)
```

```{r predict-ci}
# Point estimates and 95% credible intervals
head(data.frame(
  mean  = round(pred$mean,  3),
  lower = round(pred$lower, 3),
  upper = round(pred$upper, 3)
))
```

---

### plot()

The `plot()` method produces diagnostic and inferential graphics using the
**bayesplot** and **ggplot2** packages (both are suggested dependencies).

#### Sigma traceplot

```{r plot-trace, eval=requireNamespace("bayesplot", quietly=TRUE) && requireNamespace("ggplot2", quietly=TRUE)}
plot(fit_hs, type = "trace")
```

#### Posterior ATE distribution (causal models)

```{r plot-ate, eval=requireNamespace("bayesplot", quietly=TRUE) && requireNamespace("ggplot2", quietly=TRUE)}
plot(fit_chf, type = "ate")
```

#### CATE caterpillar plot

```{r plot-cate, eval=requireNamespace("bayesplot", quietly=TRUE) && requireNamespace("ggplot2", quietly=TRUE)}
plot(fit_chf, type = "cate")
```

#### Variable importance (Dirichlet prior)

Variable importance plots are available when `prior_type = "dirichlet"`.
For causal models with `prior_type_control = "dirichlet"` or
`prior_type_treat = "dirichlet"`, use `forest = "control"`, `"treat"`,
or `"both"`.

```{r vi-fit}
set.seed(12)
fit_dart <- ShrinkageTrees(
  y               = y_cont,
  X_train         = X_syn,
  outcome_type    = "continuous",
  prior_type      = "dirichlet",
  local_hp        = 0.1 / sqrt(5),
  number_of_trees = 5,
  N_post          = 50,
  N_burn          = 25,
  verbose         = FALSE
)
```

```{r plot-vi, eval=requireNamespace("bayesplot", quietly=TRUE) && requireNamespace("ggplot2", quietly=TRUE)}
plot(fit_dart, type = "vi", n_vi = 10)
```

---

## Multi-Chain MCMC

Running multiple independent chains improves mixing diagnostics and reduces
sensitivity to starting values. Pass `n_chains > 1` to any model-fitting
function; chains are run in parallel via `parallel::mclapply` on Unix-like
systems.

```{r multi-chain}
set.seed(13)
fit_2chain <- ShrinkageTrees(
  y               = y_cont,
  X_train         = X_syn,
  outcome_type    = "continuous",
  prior_type      = "horseshoe",
  local_hp        = 0.1 / sqrt(5),
  global_hp       = 0.1 / sqrt(5),
  number_of_trees = 5,
  N_post          = 50,
  N_burn          = 25,
  n_chains        = 2,
  verbose         = FALSE
)

cat("n_chains stored  :", fit_2chain$mcmc$n_chains, "\n")
cat("Total sigma draws:", length(fit_2chain$sigma),
    " (2 chains x 50 draws)\n")
cat("Per-chain acceptance ratios:\n")
print(round(fit_2chain$chains$acceptance_ratios, 3))
```

The same interface works for causal models.

```{r multi-chain-causal}
set.seed(14)
fit_causal_2chain <- CausalShrinkageForest(
  y                         = y_causal,
  X_train_control           = X_syn,
  X_train_treat             = X_syn,
  treatment_indicator_train = W_syn,
  outcome_type              = "continuous",
  prior_type_control        = "horseshoe",
  prior_type_treat          = "horseshoe",
  local_hp_control          = lh,
  global_hp_control         = lh,
  local_hp_treat            = lh,
  global_hp_treat           = lh,
  number_of_trees_control   = 5,
  number_of_trees_treat     = 5,
  N_post                    = 50,
  N_burn                    = 25,
  n_chains                  = 2,
  verbose                   = FALSE
)

cat("Pooled sigma draws:", length(fit_causal_2chain$sigma), "\n")
cat("Per-chain acceptance ratios (control):\n")
print(round(fit_causal_2chain$chains$acceptance_ratios_control, 3))
```

With multiple chains the traceplot shows one line per chain, and the
overlaid density plot compares the marginal posterior of $\sigma$ across
chains — both require `n_chains > 1`.

```{r plot-trace-2chain, eval=requireNamespace("bayesplot", quietly=TRUE) && requireNamespace("ggplot2", quietly=TRUE)}
plot(fit_2chain, type = "trace")
```

```{r plot-density-2chain, eval=requireNamespace("bayesplot", quietly=TRUE) && requireNamespace("ggplot2", quietly=TRUE)}
plot(fit_2chain, type = "density")
```

---

## Case Study: TCGA PAAD (Full Analysis)

The full analysis of the `pdac` dataset replicates the case study from
Jacobs, van Wieringen & van der Pas (2025). Due to the high-dimensional
covariate space (~3,000 genes) and the large MCMC settings needed for
reliable inference, the code below is provided for reference but is not
evaluated during vignette building. Pre-computed results can be reproduced
by running the `pdac_analysis` demo: `demo("pdac_analysis", package = "ShrinkageTrees")`.

### Step 1: Propensity score estimation

```{r pdac-ps, eval=FALSE}
data("pdac")

time      <- pdac$time
status    <- pdac$status
treatment <- pdac$treatment
X         <- as.matrix(pdac[, !(names(pdac) %in% c("time","status","treatment"))])

set.seed(2025)
ps_fit <- HorseTrees(
  y            = treatment,
  X_train      = X,
  outcome_type = "binary",
  k            = 0.1,
  N_post       = 5000,
  N_burn       = 5000,
  verbose      = FALSE
)

propensity <- pnorm(ps_fit$train_predictions)
```

```{r pdac-ps-overlap, eval=FALSE}
# Overlap plot
p0 <- propensity[treatment == 0]
p1 <- propensity[treatment == 1]

hist(p0, breaks = 15, col = rgb(1, 0.5, 0, 0.5), xlim = range(propensity),
     xlab = "Propensity score", main = "Propensity score overlap")
hist(p1, breaks = 15, col = rgb(0, 0.5, 0, 0.5), add = TRUE)
legend("topright", legend = c("Control", "Treated"),
       fill = c(rgb(1,0.5,0,0.5), rgb(0,0.5,0,0.5)))
```

### Step 2: Causal survival forest

```{r pdac-causal, eval=FALSE}
# Augment control matrix with propensity scores (BCF-style)
X_control <- cbind(propensity, X)

# Log-transform and centre survival times
log_time <- log(time) - mean(log(time))

set.seed(2025)
fit_pdac <- CausalHorseForest(
  y                         = log_time,
  status                    = status,
  X_train_control           = X_control,
  X_train_treat             = X,
  treatment_indicator_train = treatment,
  outcome_type              = "right-censored",
  timescale                 = "log",
  number_of_trees           = 200,
  N_post                    = 5000,
  N_burn                    = 5000,
  store_posterior_sample    = TRUE,
  verbose                   = FALSE
)
```

### Step 3: ATE and CATE estimation

```{r pdac-ate, eval=FALSE}
# Print model summary
print(fit_pdac)
smry_pdac <- summary(fit_pdac)
print(smry_pdac)

# ATE
cat("ATE posterior mean:",
    round(smry_pdac$treatment_effect$ate, 3), "\n")
cat("95% CI: [",
    round(smry_pdac$treatment_effect$ate_lower, 3), ",",
    round(smry_pdac$treatment_effect$ate_upper, 3), "]\n")
```

### Step 4: Diagnostics

```{r pdac-diag, eval=FALSE}
# Sigma convergence
plot(fit_pdac, type = "trace")

# Posterior ATE density
plot(fit_pdac, type = "ate")

# CATE caterpillar plot
plot(fit_pdac, type = "cate")
```

---

## Session Info

```{r session-info}
sessionInfo()
```

## References

Chipman, H. A., George, E. I., & McCulloch, R. E. (2010). Bayesian Additive
Regression Trees. *Annals of Applied Statistics*, 4(1), 266–298.

Hahn, P. R., Murray, J. S., & Carvalho, C. M. (2020). Bayesian Regression
Tree Models for Causal Inference: Regularization, Confounding, and
Heterogeneous Treatment Effects. *Bayesian Analysis*, 15(3), 965–1056.

Jacobs, T., van Wieringen, W. N., & van der Pas, S. L. (2025). Horseshoe
Forests for High-Dimensional Causal Survival Analysis. *arXiv preprint*
arXiv:2507.22004.

Linero, A. R. (2018). Bayesian Regression Trees for High-Dimensional
Prediction and Variable Selection. *Journal of the American Statistical
Association*, 113(522), 626–636.

Sparapani, R., Spanbauer, C., & McCulloch, R. (2021). Nonparametric Machine
Learning and Efficient Computation with Bayesian Additive Regression Trees:
The BART R Package. *Journal of Statistical Software*, 97(1), 1–66.
